Automatic discovery of cell types and microcircuitry from neural connectomics

### 2015-03-02 docs/arxiv.tex
resubmitted

%\documentclass{nature}
\documentclass{article}
\usepackage[american]{babel}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage{todonotes}
\usepackage{grffile}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{fullpage}
%\usepackage{algorithm}
%\usepackage{algorithmic}
\usepackage{amsmath}
%\usepackage{subcaption}
%\usepackage{multicol}
\usepackage{csquotes}
\usepackage{pdfpages}
\usepackage{placeins}

@@

\begin{abstract}
  Neural connectomics has begun producing massive amounts of data,
  necessitating new analysis methods to discover the biological and
  computational structure. It has long been assumed that discovering
  neuron types and their relation to microcircuitry is crucial to
  understanding neural function. Here we developed a nonparametric
  Bayesian technique that identifies neuron types and microcircuitry
  patterns in connectomics data. It combines the information
  traditionally used by biologists, including connectivity, cell body
  location and the spatial distribution of synapses.  We show that the
  approach recovers known neuron types in the retina and enables
  predictions of connectivity, better than simpler algorithms. It also
  can reveal interesting structure in the nervous system of
  c. elegans, and automatically discovers the structure of a
  microprocessor.  Our approach extracts structural meaning from
  connectomics, enabling new approaches of automatically deriving
  anatomical insights from these emerging datasets.
\end{abstract}

\section*{Introduction}
Emerging connectomics techniques \autocite{Morgan2013,Zador2012}
promise to quantify the location and connectivity of each neuron
within a tissue volume. These massive datasets will far exceed the
capacity of neuroanatomists to manually trace small circuits, thus
necessitating computational, quantitative, and automatic methods for
understanding neural circuit structure.  The impact of this kind of
high-throughput transition has been seen before -- rise of sequencing
techniques necessitated the development of novel computational methods
to understand genomic structure, ushering in  bioinformatics
as an independent discipline \autocite{Koboldt2013}.

The brain consists of multiple kinds of neurons, each of which is
hypothesized to have a specific role in the overall
computation. Neuron types differ in many ways, e.g. chemical or
morphological, but they also differ in the way they connect to one
another. In fact, the idea of well defined, type-dependent local
connectivity patterns (microcircuits) has been prominent in many
areas, from sensory (e.g. retina, \autocite{Masland2001} to processing
(e.g. neocortex \autocite{Mountcastle1997}) to movement (e.g. spinal
cord) \autocite{Grillner2005}. These sorts of repeated computing
patterns are a common feature of computing systems, even arising in
human-made computing circuits. It remains an important challenge to
develop algorithms to use anatomical data, e.g. connectomics, to
automatically back out underlying microcircuitry.

@@

The discovery of structure in probabilistic graphs is a crucial aspect
in machine learning. Commonly used algorithms include
community-based-detection methods \autocite{Girvan2002}, and
stochastic block models \autocite{Nowicki2001}.  While these
approaches can incorporate the probabilistic nature of neural
connections \autocite{Hill2012} they do not incorporate the additional
richer structure present in connectomics data -- the location of cell
bodies, the spatial distribution of synapses, and the distances
between neurons. Of particular importance is that the probability of
connections has a strong spatial component, a factor that is hard to
reconcile with many other methods. A model attempting to fully capture
the variation in the nervous system should take into account the broad
set of available features.

When it comes to neuroscience and other computing systems, we expect
patterns of connectivity much more complex than traditional motifs,
exhibiting a strong spatial dependence arising from the complex
genetic, chemical, and activity-based neural development processes.

To address these challenges, here we describe a Bayesian
nonparametric model that can discover circuit structure automatically
from connectomics data: the cell types, their spatial patterns of
interconnection, and the locations of somata and synapses. We show
that by incorporating this additional information, our model both
accurately predicts the connection as well as agrees
with human neuroanatomists as to the identification of cell types.

We primarily focus on the recently-released mouse retina connectome
\autocite{Helmstaedter2013}, but additionally examine the c. elegans
connectome \autocite{White1986}, and then ``connectome'' of a
classical microprocessor \autocite{James2010}. Comparing the cell
types discovered by the algorithms with those obtained manually by
human anatomists reveals a high degree of agreement. We present a
scalable approach to infer microcircuitry from connectomics data.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=183mm]{fig1b.ai}}
  \caption{An overview of our method. a. Input data is a connectivity
    matrix indicating whether cell $i$ (row) synapses on cell $j$
    (column), a distance matrix between cell soma, b. the depth of the
    cell soma, and the c. depth profile of the cells synaptic
    connectivity. d. Our algorithm discovers hidden cell types in this
    connectivity data by assuming all cells of a type share a
    distance-dependent connectivity profile, similar depth, and a
    similar synpatic density profile, with cells of other types.  This
    results in a clustering of the cells by those hidden
    types. d. shows the cell connectivity matrix with cells of the
    same type grouped together. e.) shows the learned probability of
    connection between our different types at various distances -- in
    this case, the cells are likely to connect when they are
    close. f. shows a connectivity profile with virtually no
    probability of connectivity. We can plot this these types g.) to
    show the relationship and spatial connections, and how probable
    various types are to connect to one another. h. The connectivity
    between our discovered types changes as a function of distance. }

@@

\section{Results}

We start with the generic assumption that cell types exist -- a cell
is a member of a single, unobserved type -- and these types affect
soma depth, distribution of synapses, as well as a cell type and
distance dependent connection probability. From this assumption we
develop a generative Bayesian model that estimates the underlying cell
types and how they connect.  We start with a model for connectivity,
the infinite stochastic block model (iSBM)\autocite{Kemp2006a,Xu2006},
which has been shown to meaningfully cluster connection graphs while
learning the number of hidden groups, or types. We extend this
approach by adding distance dependence to model salient aspects of
microcircuitry. We additionally model cell body depth as a mixture of
Gaussians and each cell’s synapse density profile as mixture of
mixtures of Gaussians.

For example, retinal ganglion cells may synapse on nearby, but not far
away, amacrine cells, with ganglion cells being superficial and having
a broad distribution of synapses.  This procedure automatically learns
the number of cell types, which cells belong to which types, along
with the distance and type dependent connectivity probabilities (see
fig~\ref{fig:overview}). We apply our algorithm to datasets from mouse
retina, c-elegans and a historical microprocessor. Anatomists classify
cells based on many features, giving us a meaningful baseline to
compare against.

Before we can use a probabilistic model to analyze real data we first
need to assess if the model works properly. We thus simulate data for
which we know the correct structure and comparing the estimated
structure based on the algorithm (see methods) with the one we used
for simulation. We find that the model does a good job of recovering
the correct number of cell types, (fig~\ref{fig:synthetic}a), the the
cell identities (fig~\ref{fig:synthetic}b), and the spatial extent of
each type (fig~\ref{fig:synthetic}c).  Indeed, it does far better than
the existing infinite stochastic block model that assumes that there
is no distance dependence to the connectivity. The model converges
relatively quickly to an estimate of the most probable values for the
cell types, which is enabled by using a combination of simulated
annealing and parallelized Markov chain Monte Carlo (see methods for
details). It appears that the model behaves the way it should and
allows application to relatively large datasets.

@@

\subsubsection{Mouse Retina}

\begin{figure}
  \centering 
  \centerline{\includegraphics{mouseretina.ai}}
  \caption{Discovering cell classes in the mouse retina connectome.
    a.) Input connectivity data for 950 cells for which soma positions
    were known. b. clustered connectivity matrix. c.) connectivity
    diagram showing our clusters, as well as the cell depth and
    anatomist-labeled cell type of each cell. Our model shows moderate
    agreement with the cell types traditionally identified by
    anatomist laminar-specific connectivity patterns. 
    d.) Connectivity between our clusters as a function
    of distance -- the cluster consisting primarily of retinal
    ganglion cells (lower left-center) exhibits the expected near and
    far connectivity. e.)  The spatial distribution of our cell types
    -- each cell type tesselates space.}
\label{fig:mouseretina}
\end{figure}

@@

As a first measure we compare link prediction accuracy across the
methods (Fig.\ref{fig:mouseretina_compare} B AUC, red).  We find that
given the dataset many techniques allow for good link-predictive
accuracy. All the methods allow decent link prediction with an AUC in
the .9 range. However, our algorithm clearly outperforms the simple
statistical models that only use connectivity.

As a second measure we compare link prediction accuracy across the
methods (Fig.\ref{fig:mouseretina_compare} B ARI, blue). We find that our
algorithm far outperforms the controls. We also find that when it is
based on more of the same information used by anatomists use then it
gets better at agreeing with these anatomists. In particular, using
connectivity, distance, synapse distribution and soma depth leads to
the highest ARI. When using the available information the algorithm
produces a good fit to human anatomist judgments.

Finally we look at the spatial extent of the discovered types both
within a layer and between layers (Fig.\ref{fig:mouseretina_compare}
C). We see that, in the absence of distance information, mere
connectivity information results in types which only span a small
region of space -- essentially “local cliques”. Incorporation of
distance information results in types which span the entire extent of
the layer. The depth variance of all models continues to be
substantially larger than that predicted by human anatomists -- future
directions of work include attempting to more strongly encode this
prior belief of laminarity.

\begin{figure}
  \centering 
  \centerline{\includegraphics{comparison.ai}}
  \caption{a. Cell-cell assignment matrix reflecting the posterior
    distribution of clusterings found by our model. Large contiguous
    black areas are cells that were considered the same type in the
    majority of samples. b. The cells and their true types sorted to
    agree with the coassignment matrix. c. A comparison of the
    predictive accuracy (Area under the curve) for hand-labeled
    anatomical data, versus inclusion of additional sources of
    information. d. ) The spatial extent (in depth and area) of the
    types identified by humans and our various algorithmic
    approaches.}

\label{fig:mouseretina_compare}
\end{figure}

@@

\subsection{C. elegans}

The roundworm Caenorhabditis elegans is a model system in
developmental neuroscience\autocite{White1986}, with the location and
connectivity of each of 302 neurons developmentally determined,
leading to early measurement of the connectome. Unlike the retina,
only the motor neurons in c. elegans exhibit regular distribution in
space -- along the body axis. Most interneurons are
concentrated in various ganglia that project throughout the entire
animal, and the sensory neurons are primarily located in a small
number of anterior ganglia.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=7in]{otherdata.ai}}
  \caption{Discovering connectivity and type in both
    \textit{c. elegans} and the MOS 6502 microprocessor.  a.) Initial
    \textit{c. elegans} cell connectivity matrix, red are
    (directional) chemical synapses, blue are electrical
    synapses. Point size is proportional to synapse count between
    those cells. b.) Clustered adjacency matrix showing identified
    clusters. c.) Inter-cluster connectivity, showing the position of
    each cell along the body axis as well as the ``true'' cell type --
    both coarse (motor, sensory, and interneuron) as well as fine
    (small text labels). d.) The clusters and connectivity versus
    distance for connections between Gate and C1, Gate and C2, and C1
    and C2 terminals on a transistor. Red and teal types have a
    terminal pulled down to ground and mostly function as
    inverters. Purple class are clocked, stateful transistors, orange
    control the ALU and yellow control the special data bus. e.) The
    spatial distribution of the transistors in each cluster.}
  \label{fig:otherdata}
\end{figure}

Using both the chemical and electrical connectivity (see methods), we
determined the underlying clusters explained by connectivity and
distance (fig~\ref{fig:otherdata}a). A superficial inspection of the results shows
clustering into groups consisting roughly homogeneously of motor
neurons, sensory neurons, and interneurons. Closer examination reveals
agreement with the classifications originally outlined by Brenner in
1986.

We identify cell types that reflect the known motor/non-motor neuron
classes, even though this system lacks the strong repeat
microcircuitry our model was designed for.  Motor-neuron types AS, DA,
and VA, all exclusively postsynaptic, are identified as a common type,
as are motor-neuron types VD and DD. Traditional types VC, DB, and VB
also mostly share a cluster. Various head motor neurons, including
types SMD and RMD, are clustered together. Interneurons with known
anatomically-distinct connectivity patterns, such as AVA (2 cells),
are clustered into pure types.The algorithm even correctly places the
single-cell types DVB and DVC by themselves.

@@

\subsection{Microprocessor}

To show the applicability of our method to other connectome-style
datasets, we obtained the spatial location and interconnectivity of
the transistors in a classic microprocessor, the MOS Technology 6502
(used in the Apple II) \autocite{James2010}. Computer architects use
common patterns of transistors when designing circuits, with each
transistor having a ``type'' in the circuit. We identified a region of
the processor with complex but known structure containing the primary
8-bit registers X, Y, and S (fig~\ref{fig:otherdata}).

>>>>

\documentclass{article}
\usepackage[american]{babel}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage[margin=true,inline=true,index=true]{fixme}
\usepackage{grffile}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{csquotes}
\usepackage{pdfpages}
\usepackage{placeins}

@@

\begin{abstract}
  Neural connectomics has begun producing massive amounts of data,
  necessitating new analysis methods to discover the biological and
  computational structure. It has long been assumed that discovering
  neuron types and their relation to microcircuitry is crucial to
  understanding neural function. Here we developed a nonparametric
  Bayesian technique that identifies neuron types and microcircuitry
  patterns in connectomics data. It combines the information
  traditionally used by biologists in a principled and probabilistically-coherent manner, including  including connectivity, cell body
  location and the spatial distribution of synapses. We show that the
  approach recovers known neuron types in the retina and enables
  predictions of connectivity, better than simpler algorithms. It also
  can reveal interesting structure in the nervous system of
  C. elegans and a historic microprocessor.
  Our approach extracts structural meaning from
  connectomics, enabling new approaches of automatically deriving
  anatomical insights from these emerging datasets.
\end{abstract}

\listoffixmes

\section*{Introduction}
Emerging connectomics techniques \autocite{Morgan2013,Zador2012}
promise to quantify the location and connectivity of each neuron
within a tissue volume. These massive datasets will far exceed the
capacity of neuroanatomists to manually trace small circuits, thus
necessitating computational, quantitative, and automatic methods for
understanding neural circuit structure.  The impact of this kind of
high-throughput transition has been seen before -- the rise of sequencing
techniques necessitated the development of novel computational methods
to understand genomic structure, ushering in bioinformatics as an
independent discipline \autocite{Koboldt2013}.

The brain consists of multiple kinds of neurons, each of which is
hypothesized to have a specific role in the overall
computation. Neuron types differ in many ways, e.g. chemical or
morphological, but they also differ in the way they connect to one
another \autocite{Seung2014}. In fact, the idea of well defined,
type-dependent local connectivity patterns (microcircuits) has a long
history \autocite{Passingham2002}, and is prominent in many areas,
from sensory (e.g. retina, \autocite{Masland2001} to processing
(e.g. neocortex \autocite{Mountcastle1997}) to movement (e.g. spinal
cord) \autocite{Grillner2005}.  These sorts of repeated computing
patterns are a common feature of computing systems, even arising in
human-made computing circuits. It remains an important challenge to
develop algorithms to use connectivity-based anatomical data
(connectomics), to automatically back out underlying microcircuitry.
already begun, focusing on

@@

The discovery of structure in probabilistic graphs is a well-known
problem in machine learning. Commonly used algorithms include
community-based-detection methods \autocite{Girvan2002}, and
stochastic block models \autocite{Nowicki2001}.  While these
approaches can incorporate the probabilistic nature of neural
connections \autocite{Hill2012} they do not incorporate the additional
richer structure present in connectomics data -- the location of cell
bodies, the spatial distribution of synapses, and the distances
between neurons. Of particular importance is that the probability of
connections has a strong spatial component, a factor that is hard to
reconcile with many other methods. A model attempting to fully capture
the variation in the nervous system should take into account the broad
set of available features.

When it comes to neuroscience and other computing systems, we expect
patterns of connectivity much more complex than traditional motifs
exhibiting a strong spatial dependence arising from the complex
genetic, chemical, and activity-based neural development processes.

To address these challenges, here we describe a Bayesian non-parametric
model that can discover circuit structure automatically from
connectomics data: the cell types, their spatial patterns of
interconnection, and the locations of somata and synapses. We show
that by incorporating this additional information, our model both
accurately predicts the connection as well as agrees with human
neuroanatomists as to the identification of cell types. We take as
inspiration previous work on identifying cell types automatically from
morphology \autocite{Guerra2011} and electrophysiology
\autocite{Druckmann2013}.

We primarily focus on the recently-released mouse retina connectome
\autocite{Helmstaedter2013}, but additionally examine the \textit{C. elegans}
connectome \autocite{White1986}.  Comparing the cell types discovered
by the algorithms with those obtained manually by human anatomists
reveals a high degree of agreement. We thus present a scalable
probabilistic approach to infer microcircuitry from connectomics data
available today and in the future.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=110mm]{fig1b.ai}}
  \caption{Deriving circuitry and cell types from connectomics data
    a. As input we take the connectivity between cells (b), the
    distance between them (c), the depth of the cell bodies (d), and
    the depth profile of the synapses (e). f. Our algorithm discovers
    hidden cell types in this connectivity data by assuming all cells
    of a type share a distance-dependent connectivity profile, similar
    depth, and a similar synaptic density profile, with cells of other
    types.  This results in a clustering of the cells by those hidden
    types. f.) shows the cell connectivity matrix with cells of the
    same type grouped together. g.) shows the learned probability of
    connection between our different types at various distances -- in
    this case, the cells are likely to connect when they are
    close. h. shows the probability of connection between two cell
    types that very rarely connect -- there's a background ``base''
    connection rate to account for errors in data, but the probability
    is very low.  i.) shows that we also recover the expected
    laminarity of types and the depth-specific synaptic
    connectivity. j.) We then plot how the connectivity between these
    types changes as a function of distance between the cell bodies to
    better understand short-range and long-range connectivity
    patterns. }

@@

\section*{Methods}

We build a structured probabilistic model which begins with the
generic notion of a cell being a member of a single type -- and these
types affect soma depth, distribution of synapses, as well as a cell
type and distance dependent connection probability. For example,
retinal ganglion cells may synapse on nearby, but not far away,
amacrine cells, with bipolar cells clearly tesselating space and
synapsing on both. In machine learning parlance, our method is
“unsupervised” -- it seeks to discover structure in data and make
predictions in the absence of training data. Rather than taking in
examples of types annotated by human neuroanatomists, we instead start
with the weakest possible assumption in an attempt to algorithmically
discover this structure. We contrast this with the supervised
approaches taken in \autocite{Guerra2011}, where there is high
confidence in the (morphologically-defined) types and then a
supervised classifier is built, as our goal here is explicitly
discovery of types.

@@

From these assumptions (priors) we develop a generative Bayesian model
that estimates the underlying cell types and how they connect. We take
as input (fig ~\ref{fig:overview}a) the connectivity matrix of cells
(fig ~\ref{fig:overview}b) , a matrix of the distance between cells
(fig ~\ref{fig:overview}c), the per-cell soma depth (fig
~\ref{fig:overview}d) and the depth profile of the cell's synapses
(fig ~\ref{fig:overview}e). We perform joint probabilistic inference
to automatically learn the number of cell types, which cells belong to
which type, their type-specific connectivity, and how connections
between types vary with distance. We also simultaneously learn the
soma depth associated with each type and the typical synaptic density
profile (fig ~\ref{fig:overview}f-h.).

We start with a model for connectivity, the infinite stochastic block
model (iSBM)\autocite{Kemp2006a,Xu2006}, which has been shown to
meaningfully cluster connection graphs while learning the number of
hidden groups, or types. We extend this approach by adding distance
dependence to model salient aspects of microcircuitry via logistic and
exponential distance-link functions.  We form a unimodial model of cell
body depth and a multimodal synapse density profile model (see
Methods for mathematical details).

As an illustrative example, consider a network with only three cell
types, labeled A, B, and C. Assume these cells are uniformly
distributed in space, and that the probability of connection between
any two cells, $c_i$ and $c_j$, depends only on their type and their
distance, according to a logistic (sigmoidal) function. Let A connect
only to nearby B and C cells, but B connect to any C regardless of
distance. This is the prior intuition our model is designed to capture.

For the basic link-distance model, we take as input a connectivity
matrix $R$ defining the connections between cell $e_i$ and $e_j$, as
well as a distance function $d(e_i, e_j)$ representing a (physical)
distance between adjacent cells.  See
the supplemental material for extension to multiple connectivity
matrices. We assume there exist an unknown number $K$ of latent
(unobserved) cell types, $k \in \{1, 2, 3, \dots, K\}$, and that each
cell $e_i$ belongs to a single cell type. We indicate a cell $e_i$ is
of type $k$ using the assignment vector $\vec(c)$, so $c_i = k$. The
observed connectivity between two cells $R(e_i, e_j)$ then depends
only on their latent type and their distance through a link function
$f(\cdot, d(e_i, e_j))$. We assume $f$ is parameterized based on the
latent type, $c_i=m$ and $c_j=n$, via a parameter $\eta_{mn}$, as well
as a set of global hyper parameters $\theta$, such that the link
function is $f(d(e_i, e_j) | \eta_{mn}, \theta)$.

We then jointly infer the posterior distribution of the
class assignment vector $\vec(c) = \{c_i\}$, the parameter matrix
$\eta_{mn}$, and the global model hyperparameters $\theta$ :

In our subsequent analysis makes use of both the full posterior
distribution over these parameters as well as the most probable, or
maximum \textit{a posteriori} (MAP), estimate.

For the retina data, we then extend the model with the additional
features indicated. Cell soma depth is modeled as a
cell-type-dependent Gaussian distribution with latent (unknown)
per-type mean and variance. Similarly, each cell has some number $N_i$
of synapses, each of which is drawn from a cell-type-specific density
profile with up to three modes.

Inference is performed via Markov-chain Monte Carlo (MCMC) via three
composable transition kernels -- one for structural, one for per-type
parameters, and one kernel for global parameters and
hyperparameters. Details of data preprocessing, inference parameters,
and runtime can be found in the Supplementary Methods section.

\subsection*{Metrics}
To evaluate the quality of the model fit, we need to use information
that quantifies aspects of the data for which we have ground truth
information. We focus on two aspects of performance. First, if the
model works well, then the probability that a pair of neurons is of
the same type should be high if the neurons actually are of the same
type. Second, the model should assign a high probability of connection
between two cells if they have a connection in the underlying data. We
term these two factors clustering accuracy and link prediction
accuracy.

To assess the accuracy of a clustering compared to that determined by
neuroanatomists, we employ three metrics -- clustering homogeneity,
clustering completness, and the adjusted Rand Index. All metrics equal
1.0 when two clusterings completely agree. Homogeneity reflects the
degree to which a found cluster or type contains only a single true
type. Completeness measures how much of a true type is contained
within a single identified type -- a completeness of 1.0 means no true
type is split into multiple subtypes. ARI is a metric that reflects both measures (See
supplemental for more information).

To assess the accuracy of the model for connections, we use link
prediction accuracy. If our model accurately captures the true
structure of the data, it should be good at predicting if a link
exists.  We thus train the model on the data with a subset of the
links marked as unobserved and thus compute our predictive
accuracy. We perform 10-way cross validation on a given dataset \autocite{Guerra2011}, learn
the resulting model, and use that model to predict the missing
synapses. Each potential link between cells is assigned a probability,
and we compute the area under the curve (AUC) for the resulting ROC
curve. An AUC of 1.0 means that we perfectly predict the presence and
absence of the missing synapses. We use link prediction accuracy to
quantify how good the model is at discovering the underlying
connectivity.

\section*{Results}
We will first establish that our algorithm works properly and try to
understand its properties using simulated data. Subsequently we will
analyze in detail a dataset on the retina. Lastly we will briefly
discuss the analysis of data from the worm \textit{c.elegans} and from
an ancient microprocessor.

\subsection*{Validation with simulated data where ground truth is known}
To validate our model, we performed a series of simulations to test if
the model can accurately recover the true underlying network structure
and cell type identity.  We thus simulate data for which we know the
correct structure and compare the estimated structure based on the
algorithm (see methods) with the one we used for simulation. We find
that the model does a good job of recovering the correct number of
cell types, (fig~\ref{fig:synthetic}a), the the cell identities
(fig~\ref{fig:synthetic}b), and the spatial extent of each type
(fig~\ref{fig:synthetic}c).  For comparison, we show the results from
instead using the infinite stochastic block model
(fig~\ref{fig:synthetic}a, b, c, black line) which assumes that only
cell type matters, and thus finds small neighborhoods of connected
nodes (instead of global connectivity patterns). This contrast shows
that while the regular block model can not correctly deal with
distance dependent connectivity, our model can. Our model converges
relatively quickly (see supplemental) to an estimate of the most
probable values for the cell types, which is enabled by using a
combination of simulated annealing and parallelized Markov chain Monte
Carlo (see methods for details). Thus our model at least is promising
to be applied to biological datasets.

\subsection*{Model Mismatch}
We next analyze how our model performs in cases where the data are
generated with assumptions different from ours.  To understand the
properties of our model, we attempt connectivity inference on four
sets of synthetic data. This helps us understand what our model would
do if the data does not obey our assumptions.

We thus generate ten sets of synthetic data from each of four existing
models.  The distance-dependent stochastic block model assumes type
depends on distance, the traditional stochastic block model has no
notion of distance, the mixed membership block model assumes type is
combinatorial, and the latent position cluster model assumes that type
is clustered-but-continuous.

If the data is sampled from our model, inference according to our
model, unsurprisingly, is good by all measures. It correctly estimates
the number of cell types; it is good at predicting connectivity (high
AUC); it agrees with human classification (Rand index); it discovers
all types, and leads to homogeneous estimates
(Fig. \ref{fig:othermodel} first row). If the data comes from a block
model without distance dependence, we see that it still does well on
all meaningful measures (Fig. \ref{fig:othermodel} second row). This
is unsurprising, as our model learns the distance dependence, even its
absence. For the mixed membership model (Fig. \ref{fig:othermodel}
third row), the model grossly overestimates the number of types, by
basically allocating a type for each combination of
memberships. Otherwise it still performs relatively well. Lastly, for
the latent position clustering model (Fig. \ref{fig:othermodel} fourth
row), the model does poorly. If type is continuous instead of discrete
then our model is basically trying to cover a continuous set with a
discrete scenario leading to rather poor performance. However, as we
do expect cell types to have a discrete biological basis we might
expect our model to do well with real data.

\begin{figure}
  \centering

  \includegraphics[width=6.0in]{synthdifferent.ai}
  \caption{Model inferences when true generating model differs from
    our distance-block-model prior. Horizontal columns show results
    with synthetic data generated according to the latent position
    cluster model, the mixed membership block model, the non-distance
    dependent stochastic block model, and the distance dependent
    stochastic block model. In all cases histograms represent
    posterior distribution over indicated metric. a.) The number of
    types found by the model, vertical dash line indicates ``true''
    type number (not applicable to mixed membership model). b.) The
    area under the ROC curve, indicating link prediction accuracy. c,
    d, e.) Clustering metrics quantifying degree of type agreement
    with known ground truth. }

  \label{fig:othermodel}
\end{figure}

\subsection*{Sensitivity to Edge Effects}

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=3.5in]{edgeeffects.pdf}}
  \caption{We generate two sets of synethetic data, one with
    spatially-dependent connectivity and one without. We measure the
    variance in the connectivity-distance plot for randomly selected
    regions of each dataset, ranging from single-cells to the entire
    volume. We see that while selecting too small of a region can 
    destroy the appearance of distance-dependent connectivity, 
    it does not create it in non-spatial data}
\label{fig:edgeeffects}
\end{figure}

Connectomic efforts so far have reconstructed only small sections of
neural tissue. Consequently, many connections to cells outside that tissue
volume will be lost. We are concerned that this selective elimination
of connectivity along the boundary might give the appearance of
distance-dependent connectivity when there is none. We thus performed
simulations to check if edge effects could destroy spatial structure
and if edge effects could introduce artificial, spurious spatial
structure. We measure the degree to which distance-dependent effects
can arise from selecting regions that are smaller than the ``scale''
of connectivity (fig ~\ref{fig:edgeeffects}). We do this by generating
two collections of synthetic datasets -- one with distance-dependent
connectivity and one without.  We then in each dataset randomly
examine contiguous circular regions with area varying from zero to the
entire volume, and empirically calculate the spatial variance in
type-dependent connectivity.  We find that, if there is no
distance-dependence, edge effects do not artificially introduce
distance dependence. However, if the section we are examining is too
small, our model can miss the distance dependence. Thus with respect
to distance-dependent connectivity inference, our model errs on the
side of caution. But we also find that for spatial extend that is
similar to the currently available datasets the effects of this are
quite limited.

\subsubsection*{Learning types and circuitry in the retina}

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=160mm]{mouseretina2.ai}}
  \caption{Discovering cell classes in the mouse retina connectome. Here
we show the maximum a posteriori (MAP) estimate for the types in the mouse
retina data. 
    a.) Input connectivity data for 950 cells for which soma positions
    were known. b. clustered connectivity matrix, each arbitrary color
    corresponds to a single type, and will be used to identify that type in the remainder
    of the plot. c.)  The spatial distribution of our cell types
    -- each cell type tessellates space. Colors correspond to those in b.)
    d.) Connectivity between our clusters as a function
    of distance -- the cluster consisting primarily of retinal
    ganglion cells (brown node on graph) exhibits the expected near and
    far connectivity.}
\label{fig:mouseretina}
\end{figure}

@@

For our model we show the receiver opererating characteristic curve
(ROC curve) (fig~\ref{fig:mouseretina_compare}d) whcih shows how the
true and false positive rates trade off. We plot the posterior
distribution of the area under this curve in
figure~\ref{fig:mouseretina_compare}e. We then plot the 
posterior distribution for cluster agreement metrics -- completeness,
homogeneity, and ari (figure~\ref{fig:mouseretina_compare}f). We see
that our model tends to over-cluster -- some of the finest-granularity of
neuroanatomist types are grouped as a single type by our model.

We compare link prediction accuracy across the
methods, including our own (Fig.\ref{fig:mouseretina_compare}g AUC, red).  We find that
given the dataset many techniques allow for good link-predictive
accuracy. All the methods allow decent link prediction with an AUC in
the 0.9 range. However, our algorithm clearly outperforms the simple
statistical models that only use connectivity.

As a second measure we compare link prediction accuracy across the
methods (Fig.\ref{fig:mouseretina_compare}g ARI, blue). We find that our
algorithm far outperforms the controls. We also find that when it is
based on more of the same information used by anatomists use then it
gets better at agreeing with these anatomists. In particular, using
connectivity, distance, synapse distribution and soma depth leads to
the highest ARI. When using the available information the algorithm
produces a good fit to human anatomist judgments.

Finally we look at the spatial extent of the discovered types both
within a layer and between layers (Fig.\ref{fig:mouseretina_compare}
h). We see that, in the absence of distance information, mere
connectivity information results in types which only span a small
region of space -- essentially “local cliques”. Incorporation of
distance information results in types which span the entire extent of
the layer. The depth variance of all models continues to be
substantially larger than that predicted by human anatomists -- future
directions of work include attempting to more strongly encode this
prior belief of laminarity.

\begin{figure}
  \centering 
  \centerline{\includegraphics{comparison.ai}}
  \caption{\textbf{Visualizing type inference uncertainty.} Our fully
    Bayesian model gives a confidence estimate (posterior probability)
    that any two given cells are of the same type. In a.) we visualize
    that cell-cell coassignment matrix, showing the probability that
    cell i is of the same type as cell j on a range from 0.0 to
    1.0. The block structure shows subsets of cells which are believed
    to all belong to the same type. For comparison, b.) shows the
    anatomist-defined type for each cell, grouped broadly into the
    coarse types identified in the previous figure. c.) for each cell
    in the matrix, c.) shows the soma depth as well as the depth
    histogram for all putative synaptic contacts. \textbf{Link vs
      cluster accuracy} d.) The posterior distribution of ROC curves
    from 10-fold cross validation when predicting connectivity, as
    well as e.) the area under the curve and f.) the type agreements
    with known neuroanatomist types. g. A comparison of the predictive
    accuracy (area under the curve) for hand-labeled anatomical data,
    versus inclusion of additional sources of information, as well as
    the clustering accuracy. Note that our model sacrifices very
    little predictive accuracy for additional clustering accuracy. By
    comparison, conventional methods fail at one or both.  d. ) The
    spatial extent (in depth and area) of the types identified by
    humans and our various algorithmic approaches.}
\label{fig:mouseretina_compare}
\end{figure}

@@

\subsection*{Recovering spatial connectivity in multiple graphs simultaneously}

Having shown our model to work on the repeating tessellated, laminar
structure of the mammalian retina, we then apply our model to a
structurally very different connectome -- the whole body of a small
roundworm: \textit{Caenorhabditis elegans} is a model system in
developmental neuroscience\autocite{White1986}, with the location and
connectivity of each of 302 neurons developmentally determined,
leading to early measurement of the connectome. Unlike the retina,
only the motor neurons in C. elegans exhibit regular distribution in
space -- along the body axis. Most interneurons are concentrated in
various ganglia that project throughout the entire animal, and the
sensory neurons are primarily located in a small number of anterior
ganglia. \textit{C. elegans} also differs from the retina in that the
measured connectome is actually two separate graphs -- one of directed
chemical synapses and another of undirected electrical synapses. As
this is a very different connectome, it allows an interesting
generalization test: how well will our model work on such a distinct
data set.

\begin{figure}
  \centering 
  \centerline{\includegraphics[width=7.0in]{celegans2.ai}}
  \caption{Discovering connectivity and type in
    \textit{C. elegans}. a.) posterior distribution on cell
    connectivity as a function of discovered type, similar to
    figure~\ref{fig:mouseretina_compare}. In b.) we plot
    neuroanatomist-derived types along with their labels. Our model
    shows high probability of motor neurons, sensory neurons, and
    various interneuron classes being of the same type. Soma positions
    along the body axis are plotted in c.) where we see that we
    cluster spatially-distributed motor neurons together, whereas head
    sensory neurons are more likely to be grouped together as well.
    d.) The ROC curves for held-out link probability for both the
    electrical synapses (gap junctions) and chemical synapses in
    c. elegans. e.) The posterior distribution of the area under the
    ROC curve for the curves in d. f.) Measurements of the agreement
    of our identified cell types compared to neuroanatomists. The high
    compleneness but low homogeneity (and corresponding low ARI)
    reflects our model's tendency to group multiple types into a
    single type.}
  \label{fig:celegans}
\end{figure}

Using both the chemical and electrical connectivity (see methods), we
determined the underlying cell types explained by connectivity and
distance (fig~\ref{fig:celegans}a). A superficial inspection of the
results shows clustering into groups consisting roughly homogeneously
of motor neurons, sensory neurons, and interneurons. Closer
examination reveals agreement with the classifications originally
outlined by White in 1986.

@@

\subsection*{Types and connectivity in artificial structures}
\begin{figure}
  \centering 
  \centerline{\includegraphics[width=6in]{mos6502.ai}}
  \caption{Discovering connectivity and type the 6502 microprocessor.
    a.) is the micrograph of the original microprocessor, with the
    region containing the registers under study highlighted. b.) Our
    graph consists of the interconnections of MOS field-effect
    transistors with three terminals, Gate, C1, and C2. Reconstruction
    technique did not permit resolution of C1 and C2 into source and
    drain. c.) The spatial distribution of the transistors in each
    cluster show a clear pattern d.) The clusters and connectivity
    versus distance for connections between Gate and C1, Gate and C2,
    and C1 and C2 terminals on a transistor. Purple and yellow types have a
    terminal pulled down to ground and mostly function as
    inverters. The blue types are clocked, stateful transistors, green
    control the ALU and orange control the special data bus (SDB).}
  \label{fig:mos6502}
\end{figure}
